Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-05-10 manuscript/Chapter2-manuscript.Rmd
repasse sur les résultats des GAMM ainsi que la discussion GAMM

Out of all five models, the two that best predicted the data, with very similar expected log pointwise densities (elpd), were the models that controlled for the prey group's speed (Table 1). Predators differed substantially in the relationship between hunting success and cumulative experience (i.e. expertise: Figure 1A-B). For instance, the success of some predators steadily increased with experience (Figure 1B, Appendix S1: Figure S1A) while others showed a decrease in success with experience (Figure 1B, Appendix S1: Figure S1B). Overall, `r toString(scales::percent(191/253))` of the population displayed an >10% increase in hunting success, `r toString(scales::percent(37/253))` displayed a >10% decrease in hunting success, and the remaining `r toString(scales::percent(25/253))` had a lesser change in success ranging between -10% and +10%.

We did not find evidence that differences among predators in the relationship between hunting success and experience was mediated by prey encounters because differences remained high even after controlling for the mean speed of the prey groups (Figure 1B). However, many individual curves appear to have shifted upwards after accounting for prey speed (Figure 1B) vs when we do not control for prey speed (Figure 1A). Moreover, in the models where we did not account for prey speed, the global trend of predator experience was nearly flat (results not shown). Yet, we see that there is an increase up to 13% in the probability of capturing all prey with experience in the population when we account for prey speed (Figure 1C).

@@

Individual predators differed slightly in their average movement speed across experience levels (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, there was strong evidence that they displayed differences in foraging specialization, with a 0.1 unit increase in standard deviation from novice to advanced, indicating that differences increased slightly (i.e. predator ID (sigma) is higher at the advanced level; Figure 2 and Appendix S1: Table S2). Moreover, some individual predators switched from a flexible to a specialized hunting tactic, and vice-versa, as they gained experience (Figure 3). Importantly, 12.6% of the predator population displayed a large change in foraging specialization/flexibility with experience (i.e. $\ge$0.2 change in standard deviation), 44.7% displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while 42.7% remained relatively stable (i.e. $\le$0.05 change in standard deviation).

@@

There was almost no correlation between IIV in predator speed and mean prey speed when predators were novice (Figure 4A). There was however a 0.17 unit increase in this correlation at the advanced stage, such that specialist hunters at this stage tended to encounter faster prey (Figure 4C). The predator IIV in speed and IIV in prey speed were moderately positively correlated (Figure 4A-B and Appendix S4: Table S1). Hence, predators that encountered groups of prey with similar speeds were more specialized in their tactic, and those that encountered groups of prey with varying speeds were more flexible. We did not find evidence that this correlation increased with experience (Figure 4C).

@@

*retravailler ici pour moins faire comme si il y avait un gros cluster. C'est surtout un continuum.*

*The maximum is attained around match 340 (i.e. advanced state). Prey speed was thus important in mediating the population and individual-level relationship between success and experience, but does appear to influence among individual differences in this relationship.*

Hunting experience is predicted to increase predator foraging specialization when resources are predictable [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, when resources vary, others predict that learning with experience should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022]. Our results show that there was no directional increase in either behavioural specialization or flexibility with experience at the population level. Instead, individual predators developed their own expertise and behavioural trajectory over successive encounters with prey; some specialized in fast-paced hunting, and others adopted a slower and more flexible tactic with experience. Specialized cursorial hunters encountered similar groups of prey while slower and flexible hunters encountered varying groups of prey, with both achieving similar success over the study period. With experience, predators increasingly matched their tactic to the speed of the prey that they encountered, which simultaneously relaxed the negative consequences of encountering faster prey. Together, our observations suggest that predators in *DBD* learn to adjust their hunting tactic to the behaviour of the prey that they encounter throughout their lifetime in the game.

@@

Empirical research shows that task proficiency often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. However, we cannot derive conclusions at the individual-level from these studies either because their models assume the same increase in expertise for each individual, or because it is unclear whether their data has repeated measures of individual success through time. Our analyses revealed that there was no population increase in expertise over time. As we mentioned before, the matchmaking pairs players using a rank-based system, and this result may thus be due to predators being paired with prey of similar skills. However, if this was solely due to the ranking system, we would have found a similar flat relationship for every predator. Instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition (Figure 1, Appendix S1: Figure S1A-S1B).

We predicted that this would occur if, by chance, predators differed in the prey that they encountered, because some may be easier and others more difficult to capture. In nature, prey can use a variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. In *DBD* we previously showed that faster movement is an effective strategy used by the prey to avoid predation [@FraserFranco.etal2022], and other studies have found that as well [@Walker.etal2005;@Kelley.Magurran2011;Martin.etal2022]. Yet, prey speed did not shape individual differences in expertise acquisition, but other antipredator tactics may be involved. For example, it is well known in *DBD* that some prey groups use mobbing to confuse the predator or prevent it from capturing an injured individual. However, like natural predator-prey systems, direct predator-mobbing in *DBD* is dangerous and requires coordinated actions that are learned through experience [@Graw.Manser2007;@Dutour.etal2016;@Pitman.etal2017;@Dutour.etal2019;@Szymkowiak2021], and was thus probably not a pervasive antipredator strategy. Other antipredator tactics known to confuse human and non-human predators during the attack sequence are unpredictable escapes and turning angles [@Jones.etal2011;@Briffa2013;@Herbert-Read.etal2017;@Moore.etal2017;@Richardson.etal2018]. While predators can adjust to unpredictable prey escapes [@Szopa-Comley.Ioannou2022], it could still be more arduous for a predator to learn and optimize its hunting success if it more often encountered prey using these tactics.

Alternatively, it is possible that individual variation in learning rates due to time delays between hunting events drove the observed differences in expertise acquisition, irrespective of the prey groups that predators encountered. For example, in *DBD*, a predator that played 300 matches in the span of six months might forget more critical information (e.g. prey escape patterns or muscle memory) than one that played 300 matches in the span of 6 days. While this has not been formally tested, a greater delay between exposure events is hypothesized to attenuate or even negate the relationship between success and experience [@Endler1991;@Wright.etal2022]. Moreover, the predators may have differed in their ability to detect prey. For example, it is widely demonstrated that predators learn visual cues such as coloration to recognize and avoid unpalatable prey [@Skelhorn.etal2016;@Ward-Fear.etal2020;@McLellan.etal2021], or habitat characteristics to locate food patches [@Ehlinger1989;@Hughes.etal1992;@Warburton2003]. @Szopa-Comley.etal2020 also found consistent among individual variation in the capacity to detect cryptic and conspicuous prey in three-spined sticklebacks (*Gasterosteus aculeatus*). Thus, individual predators should differ in their expertise if they do not learn equally how to exploit the visual cues that facilitate prey detection.

@@

@@

@@

*simplifier la 1ere phrase*

Trait-mediated predator-prey interactions are important processes involved in population cycles and should occur when predators match their phenotype to their prey [@abramsEvolutionPredatorPreyInteractions2000]. Yet, it has remained largely unknown whether this trait-matching results from predators learning how to capture their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found that individual predators displayed large differences in the development of their expertise. With experience, half of the predator population changed their behaviour by developping either a specialized or flexible tactic that was matched to the speed of their prey. Overall, both type of hunters achieved equal hunting success. Our results have potential applications for predator-prey models of trait-matching. For instance, simulation studies could incorporate individual variation in experience or learning to better predict under which ecological/evolutionary contexts specialization should be favoured over flexibility. Lastly, virtual systems are increasingly recognized among ecologists as being useful to test hypotheses on consumer-resource interactions [@Barbe.etal2020;@Beauchamp2020;@Cere.etal2021;@FraserFranco.etal2022], but also in other fields such as citizen science and conservation [@Sandbrook.etal2015;@Redpath.etal2018;@Duthie.etal2021]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

@@

**Figure 3.** Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase in specialization and (B) the greatest increase in flexibility. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of increase in either specialization or flexibility. We selected the individuals displayed in the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selecting those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialization equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal or greater than 0.28 standard deviations.

>>>>

# Open percentages table
path <- file.path(here::here("outputs", "03_outputs_model-processing"))
percent_tab <- data.table(readRDS(file.path(path, "GAMM_percentages.rds")))

Out of all five models, the two that best predicted the data, with very similar expected log pointwise densities (elpd), were the models accounting for the prey group's speed (Table 1). Predators differed substantially in the relationship between hunting success and cumulative experience (i.e. expertise: Figure 1A-B). We found evidence that these differences were in part mediated by prey speed because individuals had distinct patterns after we controlled for the mean speed of the prey groups compared to when we did not (Figure 1A-B). For instance, many individual curves shifted upwards. Precisely, the percentage of individuals that displayed a >0.5 increase in success changed from `r toString(scales::percent(percent_tab[2,]$percentage, accuracy = 0.1))` to `r toString(scales::percent(percent_tab[5,]$percentage, accuracy = 0.1))` after we accounted for prey speed (Appendix S1: Figure S1A). In this model (i.e. Figure 1B-C), `r toString(scales::percent(percent_tab[4,]$percentage, accuracy = 0.1))` of the population showed a >0.5 decrease in success (Appendix S1: Figure S1B), and the remaining `r toString(scales::percent(percent_tab[6,]$percentage, accuracy = 0.1))` displayed little to no change (between -0.5 and 0.5) in success across experience (Appendix S1: Figure S1C).

Moreover, all models where we did not account for prey speed had a global trend of predator experience that was nearly flat (results not shown). Yet, we see that there is an increase up to 13% in the probability of capturing all prey with experience in the population when we account for prey speed (Figure 1C).

@@

Individual predators differed slightly in their average movement speed across experience levels (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, there was strong evidence that they displayed differences in foraging specialization, with a 0.1 unit increase in standard deviation from novice to advanced, indicating that differences increased slightly (i.e. predator ID (sigma) is higher at the advanced level; Figure 2 and Appendix S1: Table S2). Moreover, some individual predators switched from a flexible to a specialized hunting tactic, and vice-versa, as they gained experience (Figure 3). Importantly, 12.6% of the predator population displayed a large change in foraging specialization/flexibility with experience (i.e. >0.2 change in standard deviation), 44.7% displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while 42.7% remained relatively stable (i.e. <0.05 change in standard deviation).

@@

There was almost no correlation between mean prey speed and IIV in predator speed when predators were novice (Figure 4A). There was however a 0.17 unit increase in this correlation at the advanced stage, such that advanced hunters that encountered faster prey tended to be more specialized (Figure 4C). The predator IIV in speed and IIV in prey speed were moderately positively correlated (Figure 4A-B and Appendix S4: Table S1). Hence, predators that encountered groups of prey with similar speeds were more specialized in their tactic while those that encountered groups of prey with varying speeds were more flexible. We did not however find evidence that this correlation increased with experience (Figure 4C).

@@

Hunting experience is predicted to increase predator foraging specialization when resources are predictable [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, when resources vary, others predict that learning with experience should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022]. Our results show that there was no directional increase in either behavioural specialization or flexibility with experience at the population level. Instead, individual predators developed their own expertise and behavioural trajectory over successive encounters with prey; there was a continuum of hunting styles ranging from a slower and flexible strategy to a highly specialized fast-paced strategy. Specialized cursorial hunters encountered similar groups of prey while slower and flexible hunters encountered varying groups of prey, with both achieving similar success over the study period. With experience, predators increasingly matched their foraging specialization to the speed of the prey that they encountered, but they did not achieve greater success, probably because the prey were also matching their speed to the predator. Our observations suggest that predators in *DBD* learn to adjust foraging specialization to the behaviour of the prey that they encounter throughout their lifetime in the game.

@@

Empirical research shows that task proficiency often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. Our analyses revealed a similar trend, with a 13% population increase in success with experience. However, as we mentioned before, the matchmaking pairs players using a rank-based system, and the increase in expertise could have been even greater if predators were not paired with prey of similar skill. Individual predators in *DBD* also displayed distinct patterns of expertise acquisition. We predicted that this would occur if, by chance, predators differed in the prey that they encountered, because some may be easier and others more difficult to capture. In nature, prey can use a variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. In *DBD* we previously showed that faster movement is an effective strategy used by the prey to avoid predation [@FraserFranco.etal2022], and other studies have found that as well [@Walker.etal2005;@Kelley.Magurran2011;@Martin.etal2022]. Our results confirm that prey speed influenced expertise acquisition at the individual level to some extent. This is not surprising because capturing faster prey requires costly and specialized cognitive abilities and coordination [@Kelley.Magurran2011]. However, even if we accounted for prey speed, there were still important differences in expertise acquisition, which suggests that other antipredator tactics may also be involved.

For example, two well known antipredator tactics known to confuse human and non-human predators during the attack sequence are unpredictable escapes and turning angles [@Jones.etal2011;@Herbert-Read.etal2017;@Moore.etal2017;@Richardson.etal2018]. Yet, even if predators can adjust to unpredictable prey escapes [@Szopa-Comley.Ioannou2022], it could still be arduous for an individual to learn and optimize its hunting success if it more often encountered prey using these tactics. Alternatively, it is possible that individual variation in learning rates due to time delays between hunting events drove the observed differences in expertise acquisition. For example, in *DBD*, a predator that played 300 matches in the span of six months might forget more critical information (e.g. prey escape patterns or muscle memory) than one that played 300 matches in the span of 6 days. While this has not been formally tested, a greater delay between exposure events is hypothesized to attenuate or even negate the relationship between success and experience [@Endler1991;@Wright.etal2022]. Moreover, the predators could have differed in their ability to detect prey. For instance, it is widely demonstrated that predators learn visual cues such as coloration to recognize and avoid unpalatable prey [@Skelhorn.etal2016;@Ward-Fear.etal2020;@McLellan.etal2021], or habitat characteristics to locate food patches [@Ehlinger1989;@Hughes.etal1992;@Warburton2003]. @Szopa-Comley.etal2020 also found consistent among individual variation in the capacity to detect cryptic and conspicuous prey in three-spined sticklebacks (*Gasterosteus aculeatus*). Thus, predators could differ in their expertise if they do not learn equally how to exploit the visual cues that facilitate prey detection.

@@

*je suis rendu ici pour la correction*
*m'assurer que je nuance bien et que je fais pas ressortir que l'expérience c'est full important*

@@

*ici il y a des changements importants à apporter*

@@

*arranger la conclusion*

Trait-mediated interactions that affect prey capture are the most important processes driving predator-prey dynamics [@abramsEvolutionPredatorPreyInteractions2000]. While most studies assume that predators maximize prey capture by matching their tactic to their prey [@abramsEvolutionPredatorPreyInteractions2000], it has remained largely unknown whether this results from predators learning how to hunt their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found that individual predators displayed large differences in the development of their expertise. With experience, half of the predator population changed their behaviour by developping either a specialized or flexible tactic that was matched to the speed of their prey. Overall, both type of hunters achieved equal hunting success. Our results have potential applications for predator-prey models of trait-matching. For instance, simulation studies could incorporate individual variation in experience or learning to better predict under which ecological/evolutionary contexts specialization should be favoured over flexibility. Lastly, virtual systems are increasingly recognized among ecologists as being useful to test hypotheses on consumer-resource interactions [@Barbe.etal2020;@Beauchamp2020;@Cere.etal2021;@FraserFranco.etal2022], but also in other fields such as citizen science and conservation [@Sandbrook.etal2015;@Redpath.etal2018;@Duthie.etal2021]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

@@

**Figure 3.** Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase in specialization and (B) the greatest increase in flexibility. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of increase in either specialization or flexibility. We selected the individuals displayed in the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selecting those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialization equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal or greater than 0.28 standard deviations. *check ici que les pourcentages font du sens*
